Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-05-20 text/discussion-KS-copy.qmd
discussion notes/revisions

These modest population shifts lead to substantial change in the racial/ethnic composition within and across Geomarkets over time. @fig-chicago-rq1-race illustrates the racial and ethnic composition of each Geomarket across the three census periods. In 1980, nearly all Geomarkets were predominantly White, with proportions ranging from 83% in the South and Southwest Suburbs to approximately 93% in North Shore. In contrast, the City of Chicago had a substantially more diverse population, with White residents comprising only 46% of its total population. By 2020, several Geomarkets that were previously predominantly White shifted to a more heterogeneous racial/ethnic distribution, including Chain of Lakes (59% White, 8% Black, 24% Hispanic; 7% Asian Pacific Islander), Northwest Suburbs (62% White, 3% Black, 18% Hispanic; 15% Asian Pacific Islander), Evanston and Skokie (64% White, 7% Black, 10% Hispanic; 16% Asian Pacific Islander), Western Suburbs (56% White, 9% Black, 23% Hispanic; 9% Asian Pacific Islander), and South and Southwest Suburbs (53% White, 24% Black, 18% Hispanic; 3% Asian Pacific Islander). For many Geomarkets, this balance in racial/ethnic composition was the result of large increases in the Latinx population from 1980 to 2020, which ranged from a 7 percentage point increase for the Evanston and Skokie Geomarket to a 19 percentage point increase for Western Suburbs. Some Geomarkets also experienced a substantial increase in Black populations (South and Southwest Suburbs) and in Asian populations (Northwest Suburbs and Evanston and Skokie). @fig-chicago-rq1-race also indicates that, although the North Shore and City of Chicago Geomarkets underwent some changes in racial and ethnic composition, their overall demographic patterns remained relatively the same across the three census periods. <!-- For instance, North Shore experienced some increases in Hispanic (2% in 1980 to 8% in 2020), Asian Pacific Islander (7% in 2000 to 10% in 2020), and Black (2% in 1980 to 3% in 2020) populations yet remained a predominantly White Geomarket (74% in 2020). City of Chicago also experienced relatively modest declines in its Black population (37% in 1980 to 33% in 2020), alongside increases in its Hispanic (15% in 1980 to 26% in 2020) and Asian Pacific Islander populations (4% in 2000 to 6% in 2020), further substantiating the Geomarket as the most racially/ethnic diverse in the metropolitan area. -->

@@

The discussion of homophily by @RN4975 describes not only the logic of The Market Segment model but also a broader institutional logic that characterizes the enrollment management industry. Homophily argues that there is a natural cultural match between students and colleges, suggesting colleges should find prospective students who resemble students who enrolled at their institution in the past and focus recruiting efforts on the same schools and communities that are popular with peer colleges. We suggest that homophily is the operating logic most consistent with college recruiting behavior observed in the wild. For example, @RN5023 find great overlap in the sets of private schools visited by the University of South Carolina and the University of Alabama. The study also finds great overlap in the schools visited by Villanova University and the University of Notre Dame, with a disproportionate focus on visits to Catholic schools. When we observe that students at selective private colleges tend to be rich and that out-of-state students at public research universities tend to be rich [@RN4722], this is partly because these universities devote most of their recruiting expenditures towards students from affluent schools and communities [@RN4758; @RN5023]. As @RN4975 argues, observed homophily is not the result of voluntary action; rather, homophily is programmed into algorithms that create connections.

The Market Segment Model begins with the status attainment model, that student demand for college is a function of parental education and parental income. @RN4982 [chapter 3] illustrated this argument by demonstrating a correlation between student SAT score-sending behavior and measures of class. The thesis of the Market Segment Model is that that households of specific class characteristics and college aspirations are likely to live in particular communities within metropolitan areas and that these geographic territories can be meaningfully captured by Geomarkets. Although Geomarkets appear coarse by contemporary standards, they were a landmark innovation in Geographic segmentation. The Market Segment Model recommends that colleges should identify their core student market segment (i.e., local, in-state, regional, or national), identify Geomarkets that contain large numbers of households from this market segment, and then target high schools and communities within these Geomarkets. This information is contained in the Market Segment Profile (Appendix A), a standard ouput in EPS software. When considering new Geomarkets to target, colleges should utilize the logic of homophily by identifying Geomarkets that are popular with peer colleges. This information is contained in the Institutional Profile (Appendix A), also included in EPS software.

@RN4982 state that the Market Segment Model merely formalizes the operating logic of college admissions officers. For admissions officers at selective colleges, the logic is that "good" students from "good" families are willing to attend "good" colleges that are far from home. These "good" families are found in particular parts -- Geomarkets -- of metropolitan areas. By contrast, students from other parts of the metropolitan area possess the disposition and achievement to attend their local college. The Market Segment Model formalized and commodified this logic within the EPS software to help colleges identify desirable Geomarkets and desirable high schools within those Geomarkets. @RN5007 suggest that, once commodified and quantified, the local knowledge of admissions officers becomes redundant.

Research question 1 investigates racial and socioeconomic variation between and within Geomarkets over time. We find that Geomarkets are highly correlated with race and class. \[KARINA SUMMARIZE FINDINGS IN A COUPLE SENTENCES; REVISE PARAGRAPH AS NECESSARY\] We show that in 1980, when Geomarkets were being created, that Black people tended to be highly concentrated in the poorest Geomarket in the metropolitan area. Examples include, CA7 -- City of Oakland, MA 6 -- Boston & Cambridge, OH 4 -- City of Cleveland (East), TX17 -- City of Houston (East), TX19 -- City of Dallas, CA21 -- South & South Central Los Angeles, and PA5 -- Philadelphia County. In a handful of metropolitan areas (e.g., Cleveland), Black people remain conentrated in the poorest Geomarkets. In Chicago-Land, \[KARINA -- ONE OR TWO SENTENCES\]

These findings are not surprising given the extent of race and class segregation in the US. Furthermore, the correlation between Geomarkets and class is expcted because Geomarket borders were created in service of conceptual model that views demand for higher education as a function of class. @RN4982 is written around New England, with a particular focus on Geomarkets in Connecticut and Massachusetts. @RN4982 do not discuss the intersection of student demand and race. Although unsurprising for early 1980s social science, this silence is conspicuous considering the salience of political controversy surrounding "busing" in Boston and nearby metropolitan areas.

Research question 2 analyzes the racial and socioeconomic composition of student list purchases. We examine purchases that include all Geomarkets around a metropolitan area in order to assess the potential consequences if including/excluding particular Geomarkets from the purchase. This question is motivated by the @RN4982 recommendation that selective colleges that enroll students from the regional and national market segment should target affluent Geomarkets, while community colleges and non-selective 4-year colleges should focus on middle- and working-class Geomarkets. KARINA -- ADD TEXT SUMMARIZING RESULTS.

This article adds a new perspective to existing explanations about how students are sorted into colleges. The status attainment model argues that college destination is a function of parental education and occupation [@RN1273; @RN1271; @RN1272]. The Market Segment Model is the status attainment model applied to geodemographic market research \[CITE\]. @RN5033 finds that the educational achievement of Asian American students whose parents are immigrants tend to have high educational achievement even if their parents do not, while the educational achievement of later-generation Asian Americans conforms to status attainment theory. In our analyses of prospects included in student list purchases, Asian American prospects from poor Geomarkets tended to be first-generation students while Asian American prospects from affluent Geomarkets tended to have parents with a BA. Drawing from @RN5033, one explanation for this finding is that Asian students living in low-income Geomarkets tend to have parents who immigrated to the U.S. whereas Asian students living in affluent Geomarkets are more likely to have parents who were born in the U.S.

The cultural capital model explains a process by which upper and upper-middle class families sort themselves into selective colleges by providing their children with the pedigree (academic, extracurricular) valued by selective colleges, information about how to navigate the admissions gauntlet, and social networks that provide an inside track [@RN1228; @RN1241; @RN4991; @RN1814]. This model explains how affluent households maintain a disproportionate enrollment share at selective institutions in an era of holistic admissions ostensibly designed to increase racial and class diversity [@RN4991]. Both the status attainment model and the cultural capital model are demand-side explanations for why education is a "social sieve" [@RN2377; @RN5046] that allows for a modicum of mobility while maintaining a much larger flow of intergenerational class transmission [@RN861]. @RN4722 -- who obtained parental income from federal income tax returns for every U.S. college -- show that the disproportionate enrollment share of high income families at selective private colleges -- and even at most public research universities -- is staggering. These patterns are consistent with the observation by @RN1297 [p. 241-242] that educational credentials are property that "support their holders'\ldots claims to monopolize socially and economically advantageous positions."

Supply-side explanations of credentialism and more recent scholarship on enrollment management complement the cultural capital model of sorting students to colleges. The credentialism literature recognizes that colleges have a financial incentive for educational credentials to determine the competition for socially and economically advantageous positions [@RN865; @RN861; @RN880; @RN869] and this credentials arms race favors affluent households. Scholarship on enrollment management describes which students colleges want to enroll and what colleges do to attract these students at different stages of the enrollment funnel [@RN3519; @RN4324; @RN3522; @RN4407; @RN4758; @RN3544; @RN3547; @cottom2017lower]. @RN3519 and @RN4321 describe a tacit arrangement between high school guidance counselors on the demand side and college admissions officers on the supply side. Counselors at well-resourced schools, especially private schools, are motivated to give their students a competitive advantage in admissions. Meanwhile, college admissions officers are motivated to enroll students who can afford full tuition price and are likely to donate in the future. These mutually beneficial desires are consummated by recognzing that upper and upper-middle class applicants satisfy the extracurricular needs of the college. The orchestra needs oboists. The lacrosse team needs players. Quantitative analyses of visits from colleges to high schools are consistent with @RN3519; selective private colleges and public research universities devote most of their recruiting resources on courting students from privileged schools and communities [@RN5023; @RN4758]. However, these analyses conceive of recruiting as something that is done by individual colleges.

Existing explanations of how students are sorted into colleges miss an important supply-side mechanism, third-party vendors that sort students on behalf of colleges. @RN4982 created the Market Segment Model and Geomarkets based on a snapshot of student demand from 1980. @RN4982 argues that demand for higher education is correlated with class, ignoring the historical structural barriers that produced class- and race-based inequality in 1980 student demand. In itself, the Market Segment Model is an unremarkable social science depiction. However, the College Board incscribed Geomarkets and the Market Segment Model into EPS software that told colleges which Geomarkets and high schools to target. Technologies that target target customers based on geography inevitably leverage racial and class segregation [@RN4775; @RN4849; @RN4975]. Furthermore, technologies that use the logic of predictive analytics -- identify correlates in past cases to make predictions for future cases -- are prone to "pernicious feedback loops" [@RN4849]. The structural barriers (e.g., segregation, slavery, Jim Crow) that caused historic place-based inequality in student demand are amplified because EPS recommends that colleges focus recruiting resources on localities that already have high student demand for peer colleges. Geomarkets were subsequently incorporated into the Student Search Service product. We show that that excluding low-income, non-white Geomarkets from student list purchases results in the disproportionate exclusion of first-generation, non-white students with strong test scores.

@@

The growing salience of third-party vendors warrants a reconfiguration of the organizational field salient to college access. @RN527 \[p. 148\] define organizational fields as, "those organizations that, in the aggregate, constitute a recognized area of institutional life: key suppliers, resource and product consumers, regulatory agencies, and other organizations that produce similar services or products" and that "the virtue of this unit of analysis is that it directs our attention\ldots to the totality of relevant actors." Sociological literatures salient to college access devote substantive attention to schools, colleges, families, and communities.College access scholarship from economics often analyzes the effects of district/state/federal [e.g., @RN5043] policies and the effects of information interventions [@RN4018]. These lines of research do not consider the growing role of third-party vendors, which are increasingly owned by private equity interests and offer software-as-service platforms that perform core functions for schools and colleges.[^21]

@@

Second, technological advances creates new means of identifying and serving prospects. College Board and ACT leveraged data on test-takers to enjoy a near duopoly in market for student list data, one that lasted for several decades. This business can be described as list-based lead generation based on the direct-mail model. "Free" college search engines (e.g., Cappex, Niche) yielded news sources of student list data. Another data source is college planning software purchased by high school districts and utilized by high school students and guidance counselors. The most widely-known product is Naviance, which claims to be used by more than used by more than 10 million K-12 students and by 40% ofUS high schools [@powerschool]. *Naviance* college planning software feeds into the *Intersect* recruiting platform, which allows colleges to target Naviance users while they are on the platform. *Intersect* is an example of behavioral-based targeting.

The third dynamic in the EM industry has been the growth of test-optional movement, which reduces both the number of paying test-takers and the coverage of student list products sold by College Board and ACT. ACT responded to uncertainty in their core testing business by attempting to become an edtech company \[CITE\], failing \[\], and being acquired by Nexus Capital in 2024 \[CITE\]. Over the last two decades, the testing companies developed new search filters that explicitly incorporated predictive analytics -- for example College Board's Geodemographic Segment filters \[CITE\] and ACT's predicted probability of enrollment filter \[CITE\] -- developed free college search engines, and entered the enrollment management consulting market more aggessively. College Board has weathered the test-optional movement thanks to robust revenues from the Advanced Placement product, but the decline in PSAT/SAT/PreACT/ACT test-takers has undermined their oligopoly in the market for student list data.

The fourth dynamic is the transformation from owner-operated to private-equity owned firms and erosion of the distinction between consultancy and software vendor. The 1980s and 1990s was a period of market entry in EM, when college admissions professionals or professors decided to hang up a shingle [@RN5027]. The 2010s were a period of acquisitions and concentration. After regulatory scrutiny pushed private-equity interests out of for-profit college market [@RN5022], private equity investors find value in acquiring firms that provided services to direct providers. Drawing from resource dependence theory, @list_biz describe the proliferation of horizontal acquisitions to grow market share and vertical acquisitions designed to make customers more dependent on particular recruiting products.

@@

Enroll360 exemplifies a new kind of student list product that leverages proprietary data to sell software. Historically, College Board and ACT sold prospect contact information at a price-per-prospect (e.g., \$0.50). The goal was to generate revenue from the sale of names. By contrast, Enroll360 wraps several proprietary databases of names (e.g., Cappex, Intersect) into a software-as-service platform that recruits those prospects at different stages of the enrollment funnel. Here, the goal of lists is to create demand for software, which is more expensive than names. Colleges that are interested in recruiting these prospects must buy an Enroll360 subscription. Like College Board *Student Search Service* contemporary software-as-service recruiting products incorporate search filters that enable colleges to choose which prospects they recruit. For example, reporting @markup_naviance shows how the Intersect recruiting platform allows colleges to send paid advertisements to Naviance users. Intersect customers (colleges) control which Naviance users will receive recruiting messages by filtering on filters such as "academic ability," intended majors, and wheather students used Naviance to "research competitor institutions" [@markup_naviance]. A procurement document from the University of Utah explains that the Intersect subscription is necessary to recruit the proprietary database of Naviance users: [@utah_procurement]:

> The marketing services being purchased operates within the PowerSchool Naviance platform, which is a proprietary system, that no other company has access to. There is unique group of prospective students who are only in the PowerSchool Naviance platform. We have spoken with several vendors where we can reach out to prospective students. Each vendor has their own proprietary website or database where select students can be reached.

@@

Market dynamics in the EM industry raise policy concerns about tuition prices, student choice, and competition in the market for EM consultancies may also affect competition, prices, and student choice. To the extent that colleges must buy expensive software to recruit prospects contained in an associated database, this expenditure will ultimately be passed on to students in the form of higher prices or reduced expenditure on education. Unfettered horizontal and vertical acquisitions in the EM industry causes the EM industry to become more concentrated. Smaller consultancies that do not control veritial inputs may be unable to compete with companies that have acquired proprietary pools of prospects. Fewer EM consultancies means less competition and higher consulting fees, which will be passed on to consumers. Finally, the blurring of the lines between consultancy and student list vendor suggests that large consultancies will funnel prospects only to those colleges that are willing and able to pay for subscriptions, which raises policy concerns about student choice.

@@

Regulation of the higher education industry should be informed by research that investigates the "totality of relevant actors" [@RN527, p. 148] in the organizational field. Unfortunatel, scholarship salient to education policy and regulation have ignored two, related economic transformations. First, third-party vendors offer software platforms that perform core tasks of organizations, including schools and colleges [@list_biz]. Second, private firms comprise a growing share of economic activity [@RN5037; @RN5038]. Most firms in EM and the broader edtech sector are private. However, private firms face fewer disclosure requirements than publicly traded firms. The paucity of research about third-party vendors that structure college access is substantially due to difficulty obtaining data for empirical analyses. The research community must figure out how to systematically investigate the role of private firms and investors in education, so that we can hold private interests accountable for education policy goals.

@@

We argue that sociologists and adjacent scholars should expand the domain of the education research knowledge infrastructure to include third-party vendors in enrollment management, and edtech more broadly. This is a difficult task. Student unit record data owned by district, state, and federal agencies is simultaneously granular -- can evaluate a particular policy -- and macro -- can speak to broad trends in the education sector. By contrast, obtaining granular data about third-party products in education is difficult because these data are owned by private interests that do not want researchers to interrogate their products [@RN4774; @RN4842]. We see two broad avenues for research: case studies involving primary data collection -- including this study -- which tend to have a granular focus on particular products or vendors; and secondary data analysis, which may have a more macro focus.

>>>>

These modest population shifts lead to substantial change in the racial/ethnic composition within and across Geomarkets over time. @fig-chicago-rq1-race illustrates the racial and ethnic composition of each Geomarket across the three census periods. In 1980, nearly all Geomarkets were predominantly White, with proportions ranging from 83% in the South and Southwest Suburbs to approximately 93% in North Shore. In contrast, the City of Chicago had a substantially more diverse population, with White residents comprising only 46% of its total population. By 2020, several Geomarkets that were previously predominantly White shifted to a more heterogeneous racial/ethnic distribution, including Chain of Lakes (59% White, 8% Black, 24% Hispanic; 7% Asian Pacific Islander), Northwest Suburbs (62% White, 3% Black, 18% Hispanic; 15% Asian Pacific Islander), Evanston and Skokie (64% White, 7% Black, 10% Hispanic; 16% Asian Pacific Islander), Western Suburbs (56% White, 9% Black, 23% Hispanic; 9% Asian Pacific Islander), and South and Southwest Suburbs (53% White, 24% Black, 18% Hispanic; 3% Asian Pacific Islander). For many Geomarkets, this balance in racial/ethnic composition was the result of large increases in the Hispanic population from 1980 to 2020, which ranged from a 7 percentage point increase for the Evanston and Skokie Geomarket to a 19 percentage point increase for Western Suburbs. Some Geomarkets also experienced a substantial increase in Black populations (South and Southwest Suburbs) and in Asian populations (Northwest Suburbs and Evanston and Skokie). @fig-chicago-rq1-race also indicates that, although the North Shore and City of Chicago Geomarkets underwent some changes in racial and ethnic composition, their overall demographic patterns remained relatively the same across the three census periods. <!-- For instance, North Shore experienced some increases in Hispanic (2% in 1980 to 8% in 2020), Asian Pacific Islander (7% in 2000 to 10% in 2020), and Black (2% in 1980 to 3% in 2020) populations yet remained a predominantly White Geomarket (74% in 2020). City of Chicago also experienced relatively modest declines in its Black population (37% in 1980 to 33% in 2020), alongside increases in its Hispanic (15% in 1980 to 26% in 2020) and Asian Pacific Islander populations (4% in 2000 to 6% in 2020), further substantiating the Geomarket as the most racially/ethnic diverse in the metropolitan area. -->

@@

The discussion of homophily by @RN4975 describes not only the logic of The Market Segment model but also a broader institutional logic that characterizes the enrollment management industry. Homophily argues that there is a natural cultural match between students and colleges, suggesting colleges should find prospective students who resemble students who enrolled at their institution in the past and focus recruiting efforts on the same schools and communities that are popular with peer colleges. We suggest that homophily is the operating logic most consistent with college recruiting behavior observed in the wild. For example, @RN5023 find great overlap in the sets of private schools visited by the University of South Carolina and the University of Alabama. The study also finds great overlap in the schools visited by Villanova University and the University of Notre Dame, with a disproportionate focus on visits to Catholic schools. <!--# When we observe that students at selective private colleges tend to be rich and that out-of-state students at public research universities tend to be rich [@RN4722], this is partly because these universities devote most of their recruiting expenditures towards students from affluent schools and communities [@RN4758; @RN5023]. --> As @RN4975 argues, observed homophily is not the result of voluntary action; rather, homophily is programmed into algorithms that create connections.

The Market Segment Model begins with the status attainment model, that student demand for college is a function of parental education and parental income. @RN4982 [chapter 3] illustrated this argument by demonstrating a correlation between student SAT score-sending behavior and measures of class. The thesis of the Market Segment Model is that that households of specific class characteristics and college aspirations are likely to live in particular communities within metropolitan areas and that these geographic territories can be meaningfully captured by Geomarkets. Although Geomarkets appear coarse by contemporary standards, they were a landmark innovation in Geographic segmentation. The Market Segment Model recommends that colleges should identify their core student market segment (i.e., local, in-state, regional, or national), identify Geomarkets that contain large numbers of households from this market segment, and then target high schools and communities within these Geomarkets. This information is contained in the Market Segment Profile (Appendix A), a standard ouput in EPS software. When considering new Geomarkets to target, EPS recommends colleges utilize the logic of homophily by identifying Geomarkets that are popular with peer colleges. This information is contained in the Institutional Profile (Appendix A).

<!--#  @RN4982 state that the Market Segment Model merely formalizes the operating logic of college admissions officers. For admissions officers at selective colleges, the logic is that "good" students from "good" families are willing to attend "good" colleges that are far from home. These "good" families are found in particular parts -- Geomarkets -- of metropolitan areas. By contrast, students from other parts of the metropolitan area possess the disposition and achievement to attend their local college. The Market Segment Model formalized and commodified this logic within the EPS software to help colleges identify desirable Geomarkets and desirable high schools within those Geomarkets. @RN5007 suggest that, once commodified and quantified, the local knowledge of admissions officers becomes redundant. KS: NOT SURE THIS NEEDS REPEATING-->

Research question 1 investigates racial and socioeconomic variation between and within Geomarkets over time. We find that Geomarkets are highly correlated with race and class. <!--#  [KARINA SUMMARIZE FINDINGS IN A COUPLE SENTENCES; REVISE PARAGRAPH AS NECESSARY]--> Findings illustrate that in 1980, when Geomarkets were being created, Black and Hispanic residents tended to be highly concentrated in a single, usually the poorest Geomarket in the Chicago, Dallas-Fortworth, and to some extent the Los Angeles metropolitan areas. These Geomarket patterns extend to other metropolitan areas in the country, such as CA7 -- City of Oakland, MA 6 -- Boston & Cambridge, OH 4 -- City of Cleveland (East), TX17 -- City of Houston (East), and PA5 -- Philadelphia County. By 2020, various Geomarket landscapes emerged. In Chicago, Black residents remain concentrated in the poorest Geomarket. In contrast, in Dallas-Fortworth, Black and Hispanic communities are distributed across multiple Geomarkets; however, significant socioeconomic disparities persist between Geomarkets. The Los Angeles metropolitan area presents another distinctive Geomarket landscape in which Black, Hispanic, and Asian communities across a range of socioeconomic levels are distributed across multiple Geomarkets rather than concentrated in one or two. <!--# [KARINA -- ONE OR TWO SENTENCES] -->

These findings are not surprising given the extent of race and class segregation in the US. Furthermore, the correlation between Geomarkets and class is expected because Geomarket borders were created in service of conceptual model that views demand for higher education as a function of class. @RN4982 is written around New England, with a particular focus on Geomarkets in Connecticut and Massachusetts. @RN4982 do not discuss the intersection of student demand and race. Although unsurprising for early 1980s social science, this silence is conspicuous considering the salience of political controversy surrounding "busing" in Boston and nearby metropolitan areas.<!--# KS: COULD CUT THIS IF NEEDED; MAIN TAKEAWAY IS JUST THAT RQ1 IS UNSURPRINSING EMPIRICALLY VIA SEGREGATION AND THEORETICALLY VIA ZEMSKY & OEDEL-->

Research question 2 analyzes the racial and socioeconomic composition of student list purchases. We examine purchases that include all Geomarkets in a metropolitan area in order to assess the potential consequences if including or excluding particular Geomarkets from the purchase. This question is motivated by the @RN4982 recommendation that selective colleges that enroll students from the regional and national market segment should target affluent Geomarkets, while community colleges and non-selective 4-year colleges should focus on middle- and working-class Geomarkets. Findings across metropolitan cases suggest student list orders that use Geomarket filters to exclude middle- and working-class Geomarkets - such as IL11 -- City of Chicago, TX19 -- City of Dallas, TX20 -- City of Fort Worth, CA19 -- East Los Angeles, and CA21 -- South and South Central Los Angeles, in targeting high-achieving students may still capture some Black, Hispanic, and Asian prospects. However, these prospects are far more unlikely to be first-generation students of color. <!--# KARINA -- ADD TEXT SUMMARIZING RESULTS.-->

This article adds a new perspective to existing explanations about how students are sorted into colleges. The status attainment model argues that college destination is a function of parental education and occupation [@RN1273; @RN1271; @RN1272]. The Market Segment Model is the status attainment model applied to geodemographic market research \[CITE\]. @RN5033 finds that the educational achievement of Asian American students whose parents are immigrants tend to have high educational achievement even if their parents do not, while the educational achievement of later-generation Asian Americans conforms to status attainment theory. In our analyses of prospects included in student list purchases, Asian American prospects from poor Geomarkets tended to be first-generation students while Asian American prospects from affluent Geomarkets tended to have parents with a BA. Drawing from @RN5033, one explanation for this finding is that Asian students living in low-income Geomarkets tend to have parents who immigrated to the U.S. whereas Asian students living in affluent Geomarkets are more likely to have parents who were born in the U.S. <!--# KARINA NOTES -- USE RESULTS TO SPEAK TO MAJOR THEORIES.-->

The cultural capital model explains a process by which upper and upper-middle class families sort themselves into selective colleges by providing their children with the pedigree (academic opportunities, extracurriculars) valued by selective colleges, information about how to navigate the admissions gauntlet, and social networks that provide an inside track [@RN1228; @RN1241; @RN4991; @RN1814]. This model explains how affluent households maintain a disproportionate enrollment share at selective institutions in an era of holistic admissions ostensibly designed to increase racial and class diversity [@RN4991]. Both the status attainment model and the cultural capital model are demand-side explanations for why education is a "social sieve" [@RN2377; @RN5046] that allows for a modicum of mobility while maintaining a much larger flow of intergenerational class transmission [@RN861]. @RN4722 -- who obtained parental income from federal income tax returns for every U.S. college -- show that the disproportionate enrollment share of high income families at selective private colleges -- and even at most public research universities -- is staggering. These patterns are consistent with the observation by @RN1297 [p. 241-242] that educational credentials are property that "support their holders'\ldots claims to monopolize socially and economically advantageous positions."

Supply-side explanations of credentialism and more recent scholarship on enrollment management complement the cultural capital model of sorting students to colleges. The credentialism literature recognizes that colleges have a financial incentive for educational credentials to determine the competition for socially and economically advantageous positions [@RN865; @RN861; @RN880; @RN869] and this credentials arms race favors affluent households. Scholarship on enrollment management describes which students colleges want to enroll and what colleges do to attract these students at different stages of the enrollment funnel [@RN3519; @RN4324; @RN3522; @RN4407; @RN4758; @RN3544; @RN3547; @cottom2017lower]. @RN3519 and @RN4321 describe a tacit arrangement between high school guidance counselors on the demand side and college admissions officers on the supply side. Counselors at well-resourced schools, especially private schools, are motivated to give their students a competitive advantage in admissions. Meanwhile, college admissions officers are motivated to enroll students who can afford full tuition price and are likely to donate in the future. These mutually beneficial desires are consummated by recognzing that upper and upper-middle class applicants satisfy the extracurricular needs of the college. The orchestra needs oboists. The lacrosse team needs players. Quantitative analyses of visits from colleges to high schools are consistent with @RN3519; selective private colleges and public research universities devote most of their recruiting resources on courting students from privileged schools and communities [@RN5023; @RN4758]. However, these analyses conceive of recruiting as something that is done solely by individual colleges.

Existing explanations of how students are sorted into colleges miss an important supply-side mechanism, third-party vendors that sort students on behalf of colleges. @RN4982 created the Market Segment Model and Geomarkets based on a snapshot of student demand from 1980. @RN4982 argues that demand for higher education is correlated with class, ignoring the historical structural barriers that produced class- and race-based inequality in 1980 student demand. In itself, the Market Segment Model is an unremarkable social science depiction. However, the College Board inscribed Geomarkets and the Market Segment Model into EPS software that tells colleges which Geomarkets and high schools to target. Technologies that target customers based on geography inevitably leverage racial and class segregation [@RN4775; @RN4849; @RN4975]. Furthermore, technologies that use the logic of predictive analytics -- identify correlates in past cases to make predictions for future cases -- are prone to "pernicious feedback loops" [@RN4849]. The structural barriers (e.g., segregation, slavery, Jim Crow) that caused historic place-based inequality in student demand are amplified because EPS recommends that colleges focus recruiting resources on localities that already have high student demand for peer colleges. Geomarkets were subsequently incorporated into the Student Search Service product. We show that that excluding low-income, non-white Geomarkets from student list purchases results in the disproportionate exclusion of first-generation, non-white students with strong test scores. <!--# KS Notes: THIS IS THE STRONGEST PARAGRAPH -->

@@

The growing salience of third-party vendors warrants a reconfiguration of the organizational field salient to college access. @RN527 \[p. 148\] define organizational fields as, "those organizations that, in the aggregate, constitute a recognized area of institutional life: key suppliers, resource and product consumers, regulatory agencies, and other organizations that produce similar services or products" and that "the virtue of this unit of analysis is that it directs our attention\ldots to the totality of relevant actors." Sociological literatures salient to college access devote substantive attention to schools, colleges, families, and communities. College access scholarship from economics often analyzes the effects of district, state, and federal [e.g., @RN5043] policies and the effects of information interventions [@RN4018]. These lines of research do not consider the growing role of third-party vendors, which are increasingly owned by private equity interests and offer software-as-service platforms that perform core functions for schools and colleges.[^21]

@@

Second, technological advances creates new means of identifying and serving prospects. College Board and ACT leveraged data on test-takers to enjoy a near duopoly in market for student list data, one that lasted for several decades. This business can be described as list-based lead generation based on the direct-mail model. "Free" college search engines (e.g., Cappex, Niche) yielded new sources of student list data. Another data source is college planning software purchased by high school districts and utilized by high school students and guidance counselors. The most widely-known product is Naviance, which claims to be used by more than 10 million K-12 students and by 40% of US high schools [@powerschool]. *Naviance* college planning software feeds into the *Intersect* recruiting platform, which allows colleges to target Naviance users while they are on the platform. *Intersect* is an example of behavioral-based targeting.

The third dynamic in the EM industry has been the growth of test-optional admissions movement, which reduces both the number of paying test-takers and the coverage of student list products sold by College Board and ACT. ACT responded to uncertainty in their core testing business by attempting to become an edtech company \[CITE\], failing \[\], and being acquired by Nexus Capital in 2024 \[CITE\]. Over the last two decades, the testing companies developed new search filters that explicitly incorporated predictive analytics -- for example College Board's Geodemographic Segment filters \[CITE\] and ACT's predicted probability of enrollment filter \[CITE\] -- developed free college search engines, and entered the enrollment management consulting market more aggessively. College Board has weathered the test-optional movement thanks to robust revenues from the Advanced Placement product, but the decline in PSAT/SAT/PreACT/ACT test-takers has undermined their oligopoly in the market for student list data.

The fourth dynamic is the transformation from owner-operated to private-equity owned firms and the erosion of the distinction between consultancy and software vendor. The 1980s and 1990s was a period of market entry in EM, when college admissions professionals or professors decided to hang up a shingle [@RN5027]. The 2010s were a period of acquisitions and concentration. After regulatory scrutiny pushed private-equity interests out of the for-profit college market [@RN5022], private equity investors found value in acquiring firms that provided services to direct providers. Drawing from resource dependence theory, @list_biz describe the proliferation of horizontal acquisitions to grow market share and vertical acquisitions designed to make customers more dependent on particular recruiting products.

@@

Enroll360 exemplifies a new kind of student list product that leverages proprietary data to sell software. Historically, College Board and ACT sold prospect contact information at a price-per-prospect (e.g., \$0.50 for each "name"). The goal was to generate revenue from the sale of names. By contrast, Enroll360 wraps several proprietary databases of names (e.g., Cappex, Intersect) into a software-as-service platform that recruits those prospects at different stages of the enrollment funnel. Here, the goal of lists is to create demand for software, which is more expensive than purchasing names. Colleges that are interested in recruiting these prospects must buy an Enroll360 subscription. Like College Board's *Student Search Service,* contemporary software-as-service recruiting products incorporate search filters that enable colleges to choose which prospects they recruit. For example, reporting @markup_naviance shows how the Intersect recruiting platform allows colleges to send paid advertisements to Naviance users. Intersect customers (colleges) control which Naviance users will receive recruiting messages by filtering on filters such as "academic ability," intended majors, and wheather students used Naviance to "research competitor institutions" [@markup_naviance]. <!--#  A procurement document from the University of Utah explains that the Intersect subscription is necessary to recruit the proprietary database of Naviance users: [@utah_procurement]:-->

> <!--# The marketing services being purchased operates within the PowerSchool Naviance platform, which is a proprietary system, that no other company has access to. There is unique group of prospective students who are only in the PowerSchool Naviance platform. We have spoken with several vendors where we can reach out to prospective students. Each vendor has their own proprietary website or database where select students can be reached.-->

@@

Market dynamics in the EM industry raise policy concerns about tuition prices, student choice, and competition in the market for EM consultancies that may also affect competition, prices, and student choice. To the extent that colleges must buy expensive software to recruit prospects contained in an associated database, this expenditure will ultimately be passed on to students in the form of higher prices or reduced expenditure on education. Unfettered horizontal and vertical acquisitions in the EM industry causes the EM industry to become more concentrated. Smaller consultancies that do not control vertical inputs may be unable to compete with companies that have acquired proprietary pools of prospects. Fewer EM consultancies means less competition and higher consulting fees, which will be passed on to consumers. Finally, the blurring of the lines between consultancy and student list vendor suggests that large consultancies will funnel prospects only to those colleges that are willing and able to pay for subscriptions, which raises policy concerns about student choice.

@@

Regulation of the higher education industry should be informed by research that investigates the "totality of relevant actors" [@RN527, p. 148] in the organizational field. Unfortunatel, scholarship salient to education policy and regulation have ignored two, related economic transformations. First, third-party vendors offer software platforms that perform core tasks of organizations, including schools and colleges [@list_biz]. Second, private firms comprise a growing share of economic activity [@RN5037; @RN5038]. Most firms in EM and the broader edtech sector are private. However, private firms face fewer disclosure requirements than publicly traded firms. The paucity of research about third-party vendors that structure college access is partially due to difficulty obtaining data for empirical analyses. The research community must figure out how to systematically investigate the role of private firms and investors in education, so that we can hold private interests accountable for education policy goals.

@@

We argue that sociologists and adjacent scholars should expand the domain of the education research knowledge infrastructure to include third-party vendors in enrollment management, and edtech more broadly. However, this is a difficult task. Student unit record data owned by district, state, and federal agencies is simultaneously granular -- can evaluate a particular policy -- and macro -- can speak to broad trends in the education sector. By contrast, obtaining granular data about third-party products in education is difficult because these data are owned by private interests that do not want researchers to interrogate their products [@RN4774; @RN4842]. We see two broad avenues for research: case studies involving primary data collection -- including this study -- which tend to have a granular focus on particular products or vendors; and secondary data analysis, which may have a more macro focus.
